---
title: "Help Session: In-class Hackathon and Forecasting Competition"
---

| Component       | Identification from Plot                 | Rule / Recommendation                 |
|-----------------|----------------------------------------|---------------------------------------|
| AR order (p)    | PACF significant spikes at lags 1,2…    | Number of significant spikes in PACF  |
| MA order (q)    | ACF significant spikes at lags 1,2…     | Number of significant spikes in ACF   |
| Seasonal AR (P) | PACF spike at seasonal lag (s, 2s, …)  | Number of significant seasonal spikes |
| Seasonal MA (Q) | ACF spike at seasonal lag (s, 2s, …)   | Number of significant seasonal spikes |
| Differencing (d)| Trend in series → difference once       | Use ndiffs() in R                      |
| Seasonal diff(D)| Seasonal pattern → seasonal difference | Use nsdiffs() in R                     |

| Model | ACF Behavior          | PACF Behavior         |
|-------|---------------------|---------------------|
| AR(p) | Tails off gradually  | Cuts off after lag p |
| MA(q) | Cuts off after lag q | Tails off gradually |
| ARMA(p,q) | Tails off gradually | Tails off gradually |

Additional reading

[https://chih-ling-hsu.github.io/2017/03/20/time-series](https://chih-ling-hsu.github.io/2017/03/20/time-series)


```{r, echo=FALSE, eval=FALSE}
library(forecast)

set.seed(123)

# Simulate MA(2) process
ma2 <- arima.sim(n = 100, list(ma = c(0.9, 0.5)))  # MA(2) with theta1=0.7, theta2=0.3

# Plot ACF
acf(ma2, main = "ACF of MA(2) Process")

# Plot PACF
pacf(ma2, main = "PACF of MA(2) Process")

```


## Packages

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(fable)
library(tsibble)
library(lubridate)
```

## Read Data

```{r}
library(readr)
train <- read_csv("train.csv")
#View(train)
train
```

## Convert data into tsible

```{r}
library(dplyr)

train |>
  group_by(Date) |>
  summarise(n = n()) |>
  filter(n > 1)

```


```{r}
duplicates_with_counts <- train |>
  group_by(Date) |>
  filter(n() > 1) |>       # Keep only dates that appear more than once
  mutate(count = n()) |>   # Add a column showing how many times this date appears
  ungroup() |>
  arrange(Date)

duplicates_with_counts

```


```{r}
# Correct the dates for rows 881 and 882
train <- train %>%
  mutate(Date = case_when(
    ID == 881 ~ "30.05.2025",
    ID == 882 ~ "31.05.2025",
    TRUE ~ Date
  ))
```

```{r}
train |>
  group_by(Date) |>
  filter(n() > 1) |>       # Keep only dates that appear more than once
  mutate(count = n()) |>   # Add a column showing how many times this date appears
  ungroup() |>
  arrange(Date)
```

```{r}
# --- Convert to tsibble ---
df_ts <- train |>
  mutate(Date = dmy(Date)) |>        # Convert "01.01.2023" to Date format
  as_tsibble(index = Date)

df_ts
```

## Split data to training and test

```{r}
# Find the last date
last_day <- max(df_ts$Date)

# First day of the last month
first_test_day <- floor_date(last_day, "month")

# Split
train <- df_ts %>% filter(Date < first_test_day)
test  <- df_ts %>% filter(Date >= first_test_day)

head(train)
tail(train)
head(test)
tail(test)

```

```{r}
library(feasts)
train |>
  autoplot(Value)
train |>
  ACF(Value) |>
  autoplot() +
  labs(title = "ACF Plot of Value")
train |>
  PACF(Value) |>
  autoplot() +
  labs(title = "PACF Plot of Value")
```

```{r}
train |>
  mutate(Value_diff = difference(Value)) |>  # Take first difference
  ACF(Value_diff, na_action = "omit") |>    # Compute ACF
  autoplot() +
  labs(title = "ACF Plot of Differenced Value")
```

```{r}
forecast::findfrequency(train)
```

```{r}
sarima_auto <- train |>
  model(
    auto = ARIMA(Value ~ pdq() + PDQ(period = 7))
  )

# View model summary
report(sarima_auto)
```

## Forecast for test period

```{r}
h_test <- nrow(test)

fc <- sarima_auto |>
  forecast(h = h_test)
fc
fc |> knitr::kable()
```